---
title: "Getting Started with jocf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with jocf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `jocf` package estimates conditional choice probabilities for ordered
discrete outcomes $Y \in \{1, \ldots, M\}$ given covariates $W$.  It grows a
**single** random forest whose splitting criterion jointly minimises average
estimation error across all $M$ classes simultaneously.  This is in contrast to
the published `ocf` package, which grows $M$ separate forests.

The output is an $(n \times M)$ matrix of predicted probabilities
$\hat{p}_m(w) = \hat{P}(Y = m \mid W = w)$, where rows sum to 1 by
construction.

## Basic usage

```{r fit}
library(jocf)

## Simulate ordered outcome data
set.seed(123)
n <- 200
X <- matrix(rnorm(n * 4), ncol = 4, dimnames = list(NULL, paste0("x", 1:4)))
latent <- X[, 1] - 0.5 * X[, 2] + rnorm(n)
Y <- as.integer(cut(latent, breaks = c(-Inf, -0.5, 0.5, Inf)))

## Fit a joint ordered correlation forest
fit <- jocf(Y, X, num.trees = 100)
```

The returned object is of class `"jocf"`.  In-sample predictions are stored in
`$predictions`:

```{r insample}
head(fit$predictions)

## Rows sum to 1
summary(rowSums(fit$predictions))
```

## Prediction

Use `predict()` to obtain class probabilities for new observations:

```{r predict}
X_new <- matrix(rnorm(10 * 4), ncol = 4)
preds <- predict(fit, X_new)
preds
```

## Marginal effects

`marginal_effects()` computes nonparametric average marginal effects (AME)
following Lechner (2019).  For continuous covariates, it uses finite differences;
for discrete covariates, unit changes.

```{r me}
## AME averaged over all training observations
me <- marginal_effects(fit, X)
me
```

You can evaluate at specific summary points and restrict to a subset of
covariates:

```{r me_options}
## AME evaluated at the covariate means, for covariates 1 and 3 only
me_atmean <- marginal_effects(fit, X, eval = "atmean",
                              target_covariates = c(1, 3))
me_atmean
```

If some covariates are discrete, pass their column indices via `discrete_vars`:

```{r me_discrete}
## Treat covariate 4 as discrete
me_disc <- marginal_effects(fit, X, discrete_vars = 4)
me_disc
```

## Splitting rules

`jocf()` supports two splitting rules:

- `"simple"` (default): equal weight across all $M$ classes.
- `"weighted"`: each class is weighted by the inverse of its population-level
  Bernoulli variance, giving rare classes more influence.

```{r weighted}
fit_w <- jocf(Y, X, num.trees = 100, splitting.rule = "weighted")
head(fit_w$predictions)
```

See `vignette("jocf-theory")` for the mathematical derivation.

## Parallelism

Tree growing, prediction, and marginal effects are parallelised via OpenMP.
By default, all available cores are used.  Control the number of threads with
`num.threads`:

```{r threads, eval = FALSE}
## Use 2 threads
fit2 <- jocf(Y, X, num.trees = 100, num.threads = 2)
preds2 <- predict(fit2, X_new, num.threads = 2)
me2 <- marginal_effects(fit2, X, num.threads = 2)
```
