---
title: "Getting Started with jocf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with jocf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `jocf` package estimates conditional choice probabilities and
classifications for ordered discrete outcomes $Y \in \{1, \ldots, M\}$ given
covariates $W$.  It grows a **single** random forest whose splitting criterion
jointly minimises average estimation error across all $M$ classes
simultaneously.  This is in contrast to the published `ocf` package, which
grows $M$ separate forests.

The output includes an $(n \times M)$ matrix of predicted probabilities
$\hat{p}_m(w) = \hat{P}(Y = m \mid W = w)$, where rows sum to 1 by
construction, together with two classification rules: probability-based
(argmax of averaged probabilities) and majority-vote (per-tree argmax
aggregation).  The majority-vote classifier is unique to the unified OCF and
not available to multi-forest approaches.

## Basic usage

```{r fit}
library(jocf)

## Simulate ordered outcome data
set.seed(123)
n <- 200
X <- matrix(rnorm(n * 4), ncol = 4, dimnames = list(NULL, paste0("x", 1:4)))
latent <- X[, 1] - 0.5 * X[, 2] + rnorm(n)
Y <- as.integer(cut(latent, breaks = c(-Inf, -0.5, 0.5, Inf)))

## Fit a joint ordered correlation forest
fit <- jocf(Y, X, num.trees = 100)
```

The returned object is of class `"jocf"`.  In-sample predictions are stored in
`$predictions`:

```{r insample}
head(fit$predictions)

## Rows sum to 1
summary(rowSums(fit$predictions))
```

## Classification

Both `jocf()` and `predict()` return two classification rules in
`$classification`:

- **`$prob`** — probability-based: argmax of forest-averaged probabilities.
  This targets the conditional mode and minimises 0--1 loss with respect to the
  estimated distribution.
- **`$vote`** — majority-vote: each tree votes for its leaf argmax, then the
  class receiving the most votes wins.  This rule is unique to the unified OCF
  because the original OCF has no common leaf structure across classes.

```{r classification}
## In-sample classifications
table(fit$classification$prob)
table(fit$classification$vote)
```

See `vignette("jocf-theory")` for the mathematical derivation.

## Prediction

Use `predict()` to obtain class probabilities and classifications for new
observations.  The return value is a named list with `$probabilities`
(an $(n \times M)$ matrix) and `$classification` (a list with `$prob`
and `$vote` integer vectors).

```{r predict}
X_new <- matrix(rnorm(10 * 4), ncol = 4)
preds <- predict(fit, X_new)

## Predicted probabilities (n_new x M matrix)
head(preds$probabilities)

## Probability-based classification
preds$classification$prob

## Majority-vote classification
preds$classification$vote
```

## Marginal effects

`marginal_effects()` computes nonparametric marginal effects via finite
differences.  For continuous covariates, it uses symmetric finite differences;
for discrete covariates, unit changes.  By default the training data stored in
the fitted object is used.

```{r me}
## Marginal effects averaged over all training observations
me <- marginal_effects(fit)
me
```

You can evaluate at specific summary points and restrict to a subset of
covariates via a named character vector whose names are column names and values
are `"continuous"` or `"discrete"`:

```{r me_options}
## Marginal effects evaluated at the covariate means, for x1 and x3 only
me_atmean <- marginal_effects(fit, eval = "atmean",
                              target_covariates = c(x1 = "continuous",
                                                    x3 = "continuous"))
me_atmean
```

To mark a covariate as discrete, set its value to `"discrete"`:

```{r me_discrete}
## Treat covariate x4 as discrete
me_disc <- marginal_effects(fit,
                            target_covariates = c(x4 = "discrete"))
me_disc
```

## Splitting rules

`jocf()` supports two splitting rules:

- `"simple"` (default): equal weight across all $M$ classes.
- `"weighted"`: each class is weighted by the inverse of its population-level
  Bernoulli variance, giving rare classes more influence.

```{r weighted}
fit_w <- jocf(Y, X, num.trees = 100, splitting.rule = "weighted")
head(fit_w$predictions)
```

See `vignette("jocf-theory")` for the mathematical derivation.

## Hyperparameter tuning

`jocf()` includes a built-in tuning mechanism inspired by `grf`.  Setting
`tune.parameters = "all"` tunes `mtry`, `min.node.size`, and
`sample.fraction` before growing the full forest.  The procedure grows many
small "mini-forests" at random parameter draws, computes debiased OOB error,
fits a Kriging surrogate (via `DiceKriging`), and selects the best parameters.

```{r tune, eval = requireNamespace("DiceKriging", quietly = TRUE)}
fit_tuned <- jocf(Y, X, num.trees = 100,
                  tune.parameters = "all",
                  tune.num.trees = 10,
                  tune.num.reps = 30,
                  tune.num.draws = 200)

## Selected parameters
fit_tuned$tuning.output$params

## Tuning status
fit_tuned$tuning.output$status
```

You can also tune a subset of parameters:

```{r tune_subset, eval = requireNamespace("DiceKriging", quietly = TRUE)}
fit_mtry <- jocf(Y, X, num.trees = 100,
                 tune.parameters = c("mtry"))
fit_mtry$tuning.output$params$mtry
```

## Honest forests with inference

Setting `honesty = TRUE` enables sample splitting: trees are grown on one
subsample ($S^{\text{tr}}$) and leaf predictions are computed from a held-out
honesty subsample ($S^{\text{hon}}$).  This enables valid variance estimation.

```{r honesty}
fit_h <- jocf(Y, X, num.trees = 100, honesty = TRUE, honesty.fraction = 0.5)
fit_h$honesty
```

### Predictions with standard errors

Pass `variance = TRUE` to `predict()` to obtain weight-based variance
estimates and standard errors for each predicted probability:

```{r honest_predict}
preds_h <- predict(fit_h, X_new, variance = TRUE)

## Predicted probabilities
head(preds_h$probabilities)

## Standard errors
head(preds_h$std.error)
```

### Marginal effects with standard errors

When the model is honest, `marginal_effects()` automatically returns
standard errors alongside the AME:

```{r honest_me}
me_h <- marginal_effects(fit_h)
me_h
```

See `vignette("jocf-theory")` for the mathematical details of the
weight-based inference procedure.

## Parallelism

Tree growing, prediction, and marginal effects are parallelised via OpenMP.
By default, all available cores are used.  Control the number of threads with
`num.threads`:

```{r threads, eval = FALSE}
## Use 2 threads
fit2 <- jocf(Y, X, num.trees = 100, num.threads = 2)
preds2 <- predict(fit2, X_new, num.threads = 2)
me2 <- marginal_effects(fit2, num.threads = 2)
```
